On $2$-systoles of congruence quotients of $\mathrm{SL}_3$.

### 2024-04-01 systoles-SL3.tex
Add files via upload
Changed title, added coathors, added comment about relation to Artin's primitive rooot conjecture.

\begin{document}
\title{A sample computation}	
\author{Altan Erdnigor}
\maketitle

@@

By the definition of the regulator we have
\begin{multline}
\Reg(\alpha, 2\alpha + 1) = 
\begin{vmatrix}
\ln |\alpha_1| 		& \ln |\alpha_2| \\
\ln |2 \alpha_1 + 1| 	& \ln |2 \alpha_2 + 1|
\end{vmatrix} 
\\
\approx
\begin{vmatrix}
\ln |2p + \frac{1}{2}| 	& \ln |\frac{-1}{p}| \\
\ln |4 p + 2| 	& \ln |\frac{-2}{p} + 1|
\end{vmatrix}
=
\ln(2p + \frac{1}{2})
(\ln(p - 2) - \ln(p))
+
\ln(4p + 2) \ln(p)
\\
=
\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
.\end{multline}
Therefore $ \Reg(\alpha, 2\alpha + 1) > 0 $ for all prime $p$.

Hence $\alpha, 2\alpha + 1$ are independent units.

It is left to prove that they are fundamental units in $\mathbb{Z}[\alpha]$.
By Corollary 5.9 from Conrad it is sufficient to check
\[
\frac{16 \Reg(\alpha, 2\alpha + 1) }
{(\ln(\disc_f/4))^2} 
< 2
.\] 
Substituting, we obtain
\[
\frac{16 \Reg(\alpha, 2\alpha + 1) }
{(\ln(\disc_f/4))^2} 
\approx
\frac{16 (
\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
) }
{(\ln(
(p^4 + 2 p^3 - 5 p^2 - 6 p - 23)
/4))^2} 
.\] 
Asymptotically, the latter equals
\[
\stackrel{p \to \infty}{\longrightarrow}
\frac{16 
\ln(p)^2
}
{(\ln(
p^4 ))^2} = 1
.\] 
Therefore it is $< 2$ for big enough $p$, QED.
\end{proof}
\begin{Rem}
One can do the estimate more carefully, but for now postpone it. 
\end{Rem}
\begin{Rem}
We just proved that the regulator is approximately 
\[
\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
,\] 
which is close to $\ln^2 p$ we wanted from the beginning.
\end{Rem}

\section{Centralizers}
\begin{Prop}
	\label{proposition_centralizer_in_sl}
The centralizer of $\tilde A$ in $\SL$
is generated by $A, A+\Id$.
$$Z_{\SL}(\tilde A) = \{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
\end{Prop}
\begin{proof}
Since $\tilde A$ is regular, its centralizer in $ \Mat $ is $\C \left< \Id, A, A^2 \right> $.
Now, 
\[
\C \left< \Id, A, A^2 \right> \cap \SL 
\subset \mathbb{Z} 
\left< \Id, A, A^2 \right>
.\] 
Indeed,
\begin{equation}
\Id = 
\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{pmatrix}, 
A = 
\begin{pmatrix}
0 & 0 & 1 \\
1 & 0 & p+2 \\
0 & 1 & 2p
\end{pmatrix}, 
A^2 = 
\begin{pmatrix}
0 & 1 & 2p \\
0 & p+1 & 2p^2+4p+1 \\
1 & p & 4p^2+p+2
\end{pmatrix}, 
\end{equation}
Considering the first matrix column we see that if a complex combination has integer coefficients, it is in fact integer combination.

@@

\begin{Prop}
Let $p$ be a Lucas prime.
Let $k$ be the integer part of $\log_{\phi} p$ where $\phi$ is the golden ratio.

@@

The RHS of \eqref{eq_lucas_prime_definition} being invariant under the change $\phi \to (-\phi)^{-1}$ manifests it as a symmetric polynomial in the roots of $x^2 - x - 1$, thus having a presentation
\[
\phi^k + (-\phi)^{-k} = P( \phi, (-\phi)^{-1})
,\] 
where $P$ is a \emph{universal} polynomial.
This observation justifies that the Equation \eqref{eq_lucas_prime_definition} can be taken modulo $p$ to have the form
\[
0 = \phip^k + (-\phip)^{-k} 
,\] 
which implies 
\[
1 = \phip^{4k} 
.\] 
\end{proof}

\begin{Thm}
	The index of the centralizers is bounded by $4 \log_{\phi} p$
\[
	[Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] \le 4 \log_{\phi} p
.\] 
In particular, it grows as $O(\ln p)$.
\end{Thm}
\begin{proof}
Identify $Z_{\SL}(\tilde A) \cong \mathbb{Z}/2\mathbb{Z} \oplus \mathbb{Z}^2$ as in Proposition \ref{proposition_centralizer_in_sl}.

@@

By the previous Proposition 
$Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{4k}{0}$;
Clearly, the index
\[
\mathbb{Z} \left< \binom{3}{-1}, \binom{4k}{0}\right> \subset \mathbb{Z}^2
,\] 
equals $4k \approx 4 \log_{\phi} p$ and the index 
$ [Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$
has to divide it.
\end{proof}

>>>>

\begin{document}
\title{On $2$-systoles of congruence quotients of $\mathrm{SL}_3$.}	
\author{Mikhail Belolipetsky, Altan Erdnigor, Shmuel Weinberger}
\maketitle

@@

By the definition of the regulator we have
\begin{multline}
\Reg(\alpha, 2\alpha + 1) = 
\begin{vmatrix}
\ln |\alpha_1| 		& \ln |\alpha_2| \\
\ln |2 \alpha_1 + 1| 	& \ln |2 \alpha_2 + 1|
\end{vmatrix} 
\\
\approx
\begin{vmatrix}
\ln |2p + \frac{1}{2}| 	& \ln |\frac{-1}{p}| \\
\ln |4 p + 2| 	& \ln |\frac{-2}{p} + 1|
\end{vmatrix}
=
\ln(2p + \frac{1}{2})
(\ln(p - 2) - \ln(p))
+
\ln(4p + 2) \ln(p)
\\
=
\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
.\end{multline}
Therefore, $\Reg(\alpha, 2\alpha + 1) > 0 $ for all prime $p$.

Hence $\alpha$ and $2\alpha + 1$ are independent units.

It is left to prove that they are fundamental units in $\mathbb{Z}[\alpha]$.
By Corollary 5.9 from Conrad it is sufficient to check
\[
\frac{16 \Reg(\alpha, 2\alpha + 1) }
{(\ln(\disc_f/4))^2} 
< 2
.\] 
Substituting, we obtain
\[
\frac{16 \Reg(\alpha, 2\alpha + 1) }
{(\ln(\disc_f/4))^2} 
\approx
\frac{16 (
\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
) }
{(\ln(
(p^4 + 2 p^3 - 5 p^2 - 6 p - 23)
/4))^2} 
.\] 
Asymptotically, the latter equals
\[
\stackrel{p \to \infty}{\longrightarrow}
\frac{16 
\ln(p)^2
}
{(\ln(
p^4 ))^2} = 1
.\] 
Therefore it is $< 2$ for big enough $p$.
\end{proof}
\begin{Rem}
One can do the estimate more carefully, but for now postpone it. 
\end{Rem}
\begin{Rem}
We just proved that the regulator is approximately 
\[
\ln(2 p + \frac{1}{2}) \ln(p - 2) - \ln(2 p + \frac{1}{2}) \ln(p) + \ln(4p + 2) \ln(p)
,\] 
which is close to $\ln^2 p$ we wanted from the beginning.
\end{Rem}

\section{Centralizers}
\begin{Prop}
	\label{proposition_centralizer_in_sl}
The centralizer of $\tilde A$ in $\SL$
is generated by $A, 2A+\Id$.
$$Z_{\SL}(\tilde A) = \{ \pm A^a (2 A + \Id)^b \mid a, b \in \mathbb{Z} \} .$$ 
\end{Prop}
\begin{proof}
Since $\tilde A$ is regular, its centralizer in $ \Mat $ is $\C \left< \Id, A, A^2 \right> $.
Now, 
\[
\C \left< \Id, A, A^2 \right> \cap \SL 
\subset \mathbb{Z} 
\left< \Id, A, A^2 \right>
.\] 
Indeed,
\begin{equation}
\Id = 
\begin{pmatrix}
1 & 0 & 0 \\
0 & 1 & 0 \\
0 & 0 & 1
\end{pmatrix}, 
A = 
\begin{pmatrix}
0 & 0 & 1 \\
1 & 0 & p+2 \\
0 & 1 & 2p
\end{pmatrix}, 
A^2 = 
\begin{pmatrix}
0 & 1 & 2p \\
0 & p+1 & 2p^2+4p+1 \\
1 & p & 4p^2+p+2
\end{pmatrix}, 
\end{equation}
Considering the first matrix column we see that if a complex combination has integer coefficients, it is in fact integer combination.

@@

\begin{Prop}\label{prop:lucas}
Let $p$ be a Lucas prime.
Let $k$ be the integer part of $\log_{\phi} p$ where $\phi$ is the golden ratio.

@@

The RHS of \eqref{eq_lucas_prime_definition} being invariant under the change $\phi \to (-\phi)^{-1}$ manifests it as a symmetric polynomial in the roots of $x^2 - x - 1$, thus having a presentation
\[
\phi^k + (-\phi)^{-k} = P( \phi, (-\phi)^{-1})
,\] 
where $P$ is a \emph{universal} polynomial.
This observation justifies that equation \eqref{eq_lucas_prime_definition} can be taken modulo $p$ to have the form
\[
0 = \phip^k + (-\phip)^{-k} 
,\] 
which implies 
\[
1 = \phip^{4k} 
.\] 
\end{proof}

\begin{Thm}
	If $p$ is a Lucuas prime, then the index of the centralizer is bounded by $4 \log_{\phi} p$:
\[
	[Z_{\SL}(\tilde A) : Z_{\SLp}(\tilde A)] \le 4 \log_{\phi} p
.\] 
In particular, it grows as $O(\ln p)$.
\end{Thm}
\begin{proof}
Identify $Z_{\SL}(\tilde A) \cong \mathbb{Z}/2\mathbb{Z} \oplus \mathbb{Z}^2$ as in Proposition \ref{proposition_centralizer_in_sl}.

@@

By the previous proposition 
$Z_{\SLp}(\tilde A)$ contains $\binom{3}{-1}, \binom{4k}{0}$;
Clearly, the index
\[
\mathbb{Z} \left< \binom{3}{-1}, \binom{4k}{0}\right> \subset \mathbb{Z}^2
,\] 
equals $4k \approx 4 \log_{\phi} p$ and the index 
$ [Z_{\SLp}(\tilde A) : Z_{\SL}(\tilde A)]$
has to divide it.
\end{proof}

The observation that comes next is that the sequence of Lucas primes is very sparse. It is conjectured that there are infinitely many of them but even this is not known. In any respect, it is not hard to see that the number of Lucas primes less or equal than $N$ grows at most as fast as $O(\frac{1}{\log\log N})$. From a more general point of view we are interested in the order of the generators of $Z_{\SL}(\tilde A)$ mod $p$. One of the generators can be chosen as $\tilde{A}$ and is equal one mod $p$. We are left with the other generator.  This can sometimes have very small order compared to $p$, like in the case with the Lucas primes considered in Proposition~\ref{prop:lucas}. However, in most of the cases the order will be $~p$ which implies that the index of the centralizer will be also of order $p$.
